An investor has 1000 dollars to invest in speculative stocks. The investor is considering investing $d$ dollars (a decision) in stock A and $(1000 - d)$ dollars in stock B. An investment in stock A has a $0.6$ chance of doubling in value, and a $0.4$ chance of being lost. An investment in stock B has a $0.7$ chance of doubling in value, and a $0.3$ chance of being lost. The investor's loss function for a change in fortune, $z$, is $L(\theta, z) = -\theta\log(0.0007 z + 1)$ for $-1000 \leq z \leq 1000$ and $1 \leq \theta \leq 2$. The variable $\theta$ is just an unknown parameter of the loss function in this example^[Question adapted from @Berger1985.].

a. What is the set of all possible outcomes, $\mathcal{Z}$, for a fixed $d$? (It consists of four elements.)
b. What is the expected loss, or risk function?
c. Determine the minimax decision rule $d^{\star}_{m}$.\
Hint: Plotting the risk function for fixed $\theta$ in `R`, and find its root may be helpful.

```{r id0033-risk-function-c, exercise=TRUE}



```

```{r id0033-risk-function-c-solution}
library(ggplot2)
risk <- function(d){
  - (
    0.12 * log(0.3) +
    0.18 * log(0.0014 * d + 0.3) +
    0.28 * log(1.7 - 0.0014 * d) +
    0.42 * log(1.7)
  )
  
}

ggplot(data.frame(d = c(0,1000)),aes(x = d)) + 
  stat_function(fun = risk) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_vline(xintercept = 776.96, linetype = 2) +
  ylab("r(d)")

```