---
title: "Week 0"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Week 0}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, echo = FALSE}

htmltools::includeHTML("header.html")
htmltools::includeCSS("QUTReadings.css")

```

#   Welcome to MXB341 

MXB431 is an advanced unit in mathematical statistics covering the theory of point estimation and inference using both classical and Bayesian methods.  In this class, we will learn the "how" and "why" of many statistical techniques and methods you've may have seen before. 

#### Readings and Videos

*   Instead of a traditional two-hour lecture each week, there will be
written notes or Readings posted on the unit "webpage", which you can access via Blackboard.
*   These notes are electronic documents that may be updated and revised as needed for each week. 
*   Readings contain examples with solutions (including videos).
*   It is important that you read the notes and watch the videos actively. 
*   Take notes of where you need more detail or explanation, and try on your own to solve the example problems. 
*   Most importantly, feel free to post your questions or comments about the
weeks readings and videos on the MXB341 Slack workspace.

#### On-line Consultation {#on-line-consultation .unnumbered}

*   Each week, there will be two one-hour consultation sessions.
*   Tuesdays 3-4PM consultation will take place in person in GP E-410 and on-line via Zoom. 
*   Fridays 4-5PM consultation will take place in person in GP S-520 and on line via Zoom.  
*   The notes and comments you post on Slack will serve as a guide for the teaching staff in preparing for
these consultation sessions; there will also be plenty of opportunities
for you to ask questions and interact with the teaching staff.

#### Workshops and Workshop Activities {#workshops-and-workshop-activities .unnumbered}

*   Workshops will be held Fridays from 1-3 PM, both in-person in GP G-216 and concurrently on-line via Zoom.
*   Workshops and Consultations their recordings will be available via Echo360.


#### Communications {#communications .unnumbered}

The most effective way to communicate with the teaching staff and fellow
students depends on the situation.

-   General questions about the readings, videos, workshops, assessment,
    or general (or random) questions or comments about the unit should
    be posted to the appropriate channel.

-   Posting questions to the appropriate channel ensures that all students can benefit from the response. If you have a question or
    comment, other people probably have the same questions or comment.
    Posting these on channels can help your fellow students learn and
    help the teaching staff create better resources or experiences so
    that everyone can get the most out of MXB107.

-   Email the teaching staff directly for questions about your
    circumstances or academic performance. For example, questions about
    how your assessment was marked, or other circumstances you wish to
    discuss privately with the teaching staff should be communicated via
    email.

Please note that responses to emails will be as timely as possible, but
responses to questions posted publicly may not be immediate. This delay
in reply allows you and your fellow students to discuss problems and
possible answers. Also, please note that teaching staff may have
specific times allocated for responding to students' publicly posted
questions, so that may result in a natural delay in replying to your
post.

Please maintain a respectful in any communication. It is a good idea to
address the unit teaching staff and fellow students by their
preferred name (and/or title) and use clear and precise language. Please
note that the teaching staff may work across multiple units, so your
emails should include your student id number and the unit code.

### Zoom Guidelines {#zoom-guidelines .unnumbered}

Much of the live on-line interactions will take place on Zoom; it is
important to remember that you should view these experiences the same
as classroom experiences, and you should expect the same level of
respect and consideration for Zoom sessions as you would for the
classroom. To facilitate this, here are a few guidelines:

-   When joining a Zoom session, you will be in a waiting room and
    then allowed to enter by the session host, your microphone will be
    muted. Please keep your microphone muted unless you are contributing
    to the session.

-   Be respectful of bandwidth. Please join the session with your camera turned off. Keeping your camera off helps make
    sure that everyone can benefit from the session regardless of the
    quality of their connection.

-   If you don't want to speak, please post your questions or comments so
    that everyone can read them, the host will re-read your question or
    comment aloud and respond.

-   Please realise that using technology like Zoom as a part of the learning experience is still new, and there may be issues. Please be patient and know that your constructive feedback is
    welcome and appreciated.


