---
title: "MXB341 Worksheet 08 - Bayesian Inference"
output:
  learnr::tutorial:
    theme: yeti
    includes:
    css: ["css/QUT.css","css/QUTtutorial.css"]
  progressive: true
  allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
# if problems with knitting directory:
# set Knit Directory = Document Directory

# install libraries needed for worksheet
knitr::opts_chunk$set(
  fig.align = 'center',
  collapse = TRUE,
  comment = "#>"
)

library(learnr)
library(kableExtra)
library(pander)
library(shiny)
library(MXB341)


```

```{r, echo = FALSE}

htmltools::includeHTML("header.html")
htmltools::includeCSS("./css/QUTtutorial.css")
htmltools::includeCSS("./css/QUT.css")

```


:::{.quote}
 Informed decision-making comes from a long tradition of guessing and then blaming others for inadequate results.
--- Scott Adams (Gilbert comic strip)
:::

```{r instructions-header, child='../src/instructions-header.Rmd', eval = FALSE, echo=FALSE}
  
```

# Bayesian Inference

The distinguishing aspect of Bayesian statistics is the
conceptualisation of the parameter as a probability distribution rather
than a fixed quantity. We can use this to make inference directly rather
than interpret the uncertainty in terms of the random sample via the
sampling distribution because we express the uncertainty about the
parameter as a probability distribution. We can use the posterior
distribution to make direct probabilistic statements about the parameter
in some ways (confidence or credible intervals). While this makes
interpreting interval estimates more straightforward, it is more
computationally demanding as it requires choosing an appropriate prior
and evaluating the posterior distribution. In some cases, there are no
analogous procedures to classical or frequentist approaches
(i.e. hypothesis testing). Lastly, it is important to recognise and
understand the influence of the parameter's prior distribution on the
results of analyses.

Thus we will examine three approaches to inference in the Bayesian
paradigm: interval estimation, hypothesis testing, and decision theory.
Each of these approaches has its analogue with classical techniques, and
we will explore these contrasts and comparisons as part of exploring
these concepts.

##    Credible Intervals

Interval estimation in the Bayesian context uses the posterior
distribution to directly compute an interval-based on a specified
probability

::: {.boxed}
**Credible Intervals**\

Given the posterior distribution $\pi(\theta|\mathbf{x})$ the $1-\alpha$
***credible set*** is
$$C(\theta)=\left\{\theta:Pr\left(\theta\in C(\theta)\right)=1-\alpha\right\}$$
For univariate cases this can be defined by the interval $[a,b]$ $$
\int_a^b\pi(\theta|\mathbf{y})d\theta=1-\alpha
$$ thus $$
C(\theta)=\{\theta:\theta\in[a,b]\}.
$$ As in the case of confidence intervals, there are limited analytical
tools available to minimise $b-a$ for a given $1-\alpha$. In many
instances, numerical techniques are necessary.

A set $C^*(\theta)=[a,b]$ that minimises $b-a$ for a given $1-\alpha$ $$
C(\theta)=\min_{b-a}\int_{a}^b\pi(\theta|\mathbf{y})d\theta=1-\alpha
$$ is the ***highest posterior density*** credible interval.
:::


##    Hypothesis Testing

For hypotheses of the form 
$$
H_0:\theta\in \theta_0\quad\mbox{vs}\quad H_A:\theta\in \theta_0^c
$$ 
the posterior probabilities are equivalent to testing the hypotheses
$$
\begin{align}
Pr(\theta\in\theta_0|\mathbf{y})&\equiv Pr(\mbox{H_0 is True}|\mathbf{y})\\
Pr(\theta\in\theta_0^c|\mathbf{y})&\equiv Pr(\mbox{H_0 is False}|\mathbf{y}).
\end{align}
$$

But in the case of the point-null hypothesis 
$$
H_0:\theta=\theta_0\quad\mbox{vs}\quad H_A:\theta\neq \theta_0
$$ 
$Pr(\theta=\theta_0)=0$ which prevents meaningful hypothesis testing.
(The exception being when the parameter space is discrete, but this is a
rare occurrence.)

## Bayes Factors
Bayes factors are one alternative to classical hypothesis testing. If there are two competing models, $M_{1}$ and $M_{2}$, we compare the marginal likelihood of the hypotheses they represent. The marginal likelihood is the probability of observing the data for a given model, i.e. $P(\text{data}) = P(\boldsymbol{x})$  (this is the normalising constant in Bayes rule above!). Since we have two different models we add the conditioning:
$$
Pr(\mathbf{x}\vert~M_{i}) = Pr(\boldsymbol{x}\vert M_{i}) = \int_{\theta_{i}} f(\boldsymbol{x}\vert\theta_{i}, M_{i})\pi(\theta_{i}\vert M_{i}) \text{ d}\theta_{i}
$$
but can also be written as
$$
Pr(\boldsymbol{x}\vert M_{i}) =\frac{Pr(M_{i} \vert \boldsymbol{x})Pr(\boldsymbol{x})}{Pr(M_{i})}
$$
using Bayes rule. The Bayes factor comparing these two models is
$$
BF_{12} = \frac{Pr(\boldsymbol{x}~\vert~M_{1})}{Pr(\boldsymbol{x}\vert M_{2})} = \frac{Pr(M_{1} \vert \boldsymbol{x}) Pr(M_{2})}{Pr(M_{2}\vert\boldsymbol{x}) Pr(M_{1})}
$$
the interpretation of $BF_{12}$ is given by tables, but there is more than one that can be used.

##    Bayesian Decision Theory



## Theory questions


### Question 1

:::{.boxed}

```{r theory-question-1, child='../question-bank/id0030-bayes-lomax.Rmd', eval = TRUE}
  
```
:::

<button class="solution">
Solution
</button>
:::{.panel}
```{r theory-solution-1, child='../solution-bank/id0030-bayes-lomax.Rmd', eval = TRUE}
  
```
:::


### Question 2

:::{.boxed}

```{r theory-question-2, child='../question-bank/id0048-bayes-credible-interval.Rmd', eval = TRUE}
  
```
:::

<button class="solution">
Solution
</button>
:::{.panel}
```{r theory-solution-2, child='../solution-bank/id0048-bayes-credible-interval.Rmd', eval = TRUE}

```
:::


### Question 3

:::{.boxed}

```{r theory-question-3, child='../question-bank/id0035-decision-bayes-risk-sim-prac.Rmd', eval = TRUE}
  
```
:::

<button class="solution">
Solution
</button>
:::{.panel}
```{r theory-solution-3, child='../solution-bank/id0035-decision-bayes-risk-sim-prac.Rmd', eval = TRUE}

```
:::

## Practical questions


### Question 4

:::{.boxed}

```{r prac-question-4, child='../question-bank/id0032-bayes-factor-coco-prac.Rmd', eval = TRUE}
  
```

:::

<button class="solution">
Solution
</button>
:::{.panel}
```{r prac-solution-4, child='../solution-bank/id0032-bayes-factor-coco-prac.Rmd', eval = TRUE}
  
```
:::

```{r,echo=FALSE}
htmltools::includeHTML("footer.html")
```
