
*Solution*: 

a. Under loss function optimality the best estimator for $\theta$ is 
$$
\underset{\theta\in\Theta}{\operatorname{argmin}}(d-\theta)^4-(d-\theta)^3+(d-\theta)^2+(d-\theta)
$$
We can find this analytically by letting 
$$
z=(d-\theta) 
$$
setting the derivative of 
$$
z^4-z^3+z^2+z
$$
to $0$ and solving for the roots of the cubic equations, then verifying that it is a minima.  Or we can do this numerically and graphically

```{r pq1, exercise = TRUE}

##  Define a function for the polynomial in z

## Plot the function over a range of values (positive and negative)

## Use optim() to find the minima


```

```{r pq1-solution}

##  Define a function for the polynomial in z

f<-function(z)
{
  z^4-z^3+z^2+z
}

## Plot the function over a range of values (positive and negative)

x<-seq(-1,1,len=100)

df<-tibble(x=x,y=f(x))
ggplot(df,aes(x=x,y=y))+
  geom_line()+
  xlab(expression("z="*(d-theta)))+
  ylab(expression("quartic loss function: "*z^4-z^3+z^2+z))

## Use optimise() to find the minima

res<-optimise(f,c(-1,1))




```

b. 

The log likelihood with respect to 

```{r pq2, exercise = TRUE}

##  Define functions for the MLE and the quartic losses

##  Plot in terms of d-\theta

```

```{r pq2-solution}

##  Define functions for the MLE and the quartic losses

##  See above, for the Poisson MLE you can just use -dpois()

##  Plot in terms of d-\theta

x<-seq(-1,1,len=100)

X<-rep(x,2)
Y<-c(f(x),-dpois(1,1+x,log=TRUE))

df<-tibble(x=X,y=Y,name=c(rep("quartic",100),rep("MLE",100)))
ggplot(df,aes(x=x,y=y,color=name))+
  geom_line()+
  xlab(expression("z="*(d-theta)))+
  ylab(expression("loss function"))



```

c.    The quartic loss function is biased, it prefers to underestimate $\theta$, this is an example of a shrinkage estiamtor.  



