Given 
$$
y=-3.48, -2.82, -1.96,  0.48, -2.59,  2.48,  3.06,  4.05,  3.80,  6.12,  5.98
$$
and
$$
x = -5, -4, -3, -2, -1,  0,  1,  2,  3,  4,  5
$$
Implement a Gibbs sampler for the model in question 2 (theory) using `R`. 

(a) Write a one-dimensional MH sampler for the conditional distribution $(\sigma^2~\vert~\beta,\alpha, \boldsymbol{y})$ assuming that the prior 
$$
\pi(\sigma^2)=\frac{1}{(1+\sigma^2)^2}
$$ 
Check that it works.
(b) Write a Gibbs sampler for $(\beta,\alpha~\vert~\sigma^2, \boldsymbol{y})$. Check that it works as expected.
(c) Write a MH-within-Gibbs sampler for the full distribution using the solutions to (a) and (b). The Gibbs step for $\sigma^2$ should simply use the MH step in (a).

Finally, comment on the performance of these algorithms

**Solution:**\

a.  

```{r part-a}
x<-c(-5,-4,-3,-2,-1,0,1,2,3,4,5)
y<-c( -3.48, -2.82, -1.96,  0.48, -2.59,  2.48,  3.06,  4.05,  3.80,  6.12,  5.98)
xbar<-mean(x)

log_like<-function(alpha,beta,sigma2)
{
  dnorm(y, mean = alpha+beta*(x-xbar),sd=sqrt(sigma2),log = TRUE)%>%sum()
}

##  Fix beta=alpha=1

beta<-1
alpha<-1

n<-1000

s2<-numeric(length = n)

s2[1]<-1

for(i in 2:n)
{
  sc<-rexp(1,rate = 1/s2[i-1])
  num<-log_like(alpha,beta,sc)-2*log(sc+1)-s2[i-1]/sc
  den<-log_like(alpha,beta,s2[i-1])-2*log(s2[i-1]+1)-sc/s2[i-1]
  a<-(num-den)%>%exp()
  u<-runif(1)
  if(u<a)
  {
    s2[i]<-sc
  }else
    s2[i]<-s2[i-1]
}

```

b. 

```{r part-b}

x<-c(-5,-4,-3,-2,-1,0,1,2,3,4,5)
y<-c( -3.48, -2.82, -1.96,  0.48, -2.59,  2.48,  3.06,  4.05,  3.80,  6.12,  5.98)
xbar<-mean(x)

log_like<-function(alpha,beta,sigma2)
{
  dnorm(y, mean = alpha+beta*(x-xbar),sd=sqrt(sigma2),log = TRUE)%>%sum()
}

n<-1000

##  Fix s2<-1

s2<-1

beta<-numeric(length = n)
alpha<-numeric(length = n)
beta[1]<-0
alpha[1]<-0

for(i in 2:n)
{
  bc<-rnorm(1,mean=beta[i-1],sd=0.3)
  ac<-rnorm(1,mean=alpha[i-1],sd=0.3)
  num<-log_like(ac,bc,s2)
  den<-log_like(alpha[i-1],beta[i-1],s2)
  a<-(num-den)%>%exp()
  u<-runif(1)
  if(u<a)
  {
    alpha[i]<-ac
    beta[i]<-bc
  }else
    {
    alpha[i]<-alpha[i-1]
    beta[i]<-beta[i-1]
    }  
}


```

c.    

```{r part-c}

x<-c(-5,-4,-3,-2,-1,0,1,2,3,4,5)
y<-c( -3.48, -2.82, -1.96,  0.48, -2.59,  2.48,  3.06,  4.05,  3.80,  6.12,  5.98)
xbar<-mean(x)

log_like<-function(alpha,beta,sigma2)
{
  dnorm(y, mean = alpha+beta*(x-xbar),sd=sqrt(sigma2),log = TRUE)%>%sum()
}

n<-1000

beta<-numeric(length = n)
alpha<-numeric(length = n)
s2<-numeric(length = n)
beta[1]<-0
alpha[1]<-0
s2[1]<-1

for(i in 2:n)
{
  bc<-rnorm(1,mean=beta[i-1],sd=0.05)
  ac<-rnorm(1,mean=alpha[i-1],sd=0.05)
  num<-log_like(ac,bc,s2[i-1])
  den<-log_like(alpha[i-1],beta[i-1],s2[i-1])
  a<-(num-den)%>%exp()
  u<-runif(1)
  if(u<a)
  {
    alpha[i]<-ac
    beta[i]<-bc
  }else
    {
    alpha[i]<-alpha[i-1]
    beta[i]<-beta[i-1]
    }  
  sc<-rexp(1,rate = 1/s2[i-1])
  num<-log_like(alpha[i],beta[i],sc)-2*log(sc+1)-s2[i-1]/sc
  den<-log_like(alpha[i],beta[i],s2[i-1])-2*log(s2[i-1]+1)-sc/s2[i-1]
  a<-(num-den)%>%exp()
  u<-runif(1)
  if(u<a)
  {
    s2[i]<-sc
  }else
    s2[i]<-s2[i-1]
}

