
***
*Solution*: 

(a)

$$b = \sup_{x}\sqrt{h(x)} = \sup_{x} \exp\{-x/2\} = 1$$
when $x = 0$.

For values $b$ and $c$, let $g(x) =  x \exp\{-x/2\}$

$$\frac{\partial g}{\partial x} = \exp\{-x/2\} -\frac{x}{2} \exp\{-x/2\}$$

$\frac{\partial g}{\partial x} = 0$ when $x = 2$, which can be confirmed to be a maximum (and hence suprememum). Therefore, $d = \sup_{x} x \sqrt{h(x)}$.

$$d = g(2) = 2 \exp\{-2/2\} = \frac{2}{e}$$

The infimum is therefore at $x=0$ or as $x \rightarrow \infty$.

$g(0) = 0$ and $\lim_{x \rightarrow \infty} g(x) = 0$ also, therefore $d = \sup_{x} x \sqrt{h(x)} = 0$.

(b)

From (a), the bounds for the uniform random variables are:

$0 \leq u_{1} \leq 1$ and $0 \leq u_{2} \leq \frac{2}{e}$

```{r id0041-sol-b}

exp_sampler <- function(){
  
  max_attempts <- 100
  
  h <- function(x) exp(-x)
  b <- 1
  c <- 0
  d <- 2/exp(1)
    
  for(i in 1:max_attempts){
    u1 <- runif(1, min = 0, max = b)
    u2 <- runif(1, min = c, max = d)
    
    x <- u2/u1
    accept <- u1 < sqrt(h(x))
    
    if(accept) return(x)
  }
  
}

x <- replicate(exp_sampler(), n = 1000)
y <- rexp(1000, rate = 1)

mean(x);mean(y)
var(x);var(y)

ggplot(data.frame(x=x,y=y)) + 
  geom_density(aes(x=x), colour = "red") + 
  geom_density(aes(x=y), colour = "black")

```

(c) Yes, if $x \sim \text{Exp}(1)$ then $z = \frac{x}{\lambda}$ has distribution $z \sim \text{Exp}(\lambda)$.

```{r id0041-sol-c}

exp_sampler_lambda <- function(lambda){
  
  exp_sampler() / lambda
  
}

x <- replicate(exp_sampler_lambda(2), n = 1000)
y <- rexp(1000, rate = 2)

mean(x);mean(y)
var(x);var(y)

ggplot(data.frame(x=x,y=y)) + 
  geom_density(aes(x=x), colour = "red") + 
  geom_density(aes(x=y), colour = "black")

```

***