Given 
```{r echo=FALSE}
set.seed(14061972)

x<-runif(20,-1,1)

lambda<-exp(3.1*x-1.7)

y<-rpois(20,lambda=lambda)

```
$y=$ `r round(y,2)`  and \
$x =$ `r round(x,2)`

(a) Fit a model assuming that $y_i\sim Pois(\lambda_i)$, $\log(\lambda_i)=\beta_0+\beta_1x_i$.
(b) Plot the cummulative mean of the parameters and calculate the BGR for three chains.

**Solution:**\
Write a function to evaluate the model:
```{r}
library(mvtnorm)

x<-c(0.59, -0.67, 0.7, -0.86, -0.99, 0.92, 0.57, -0.04, -0.57, -0.74, 0.98, -0.81, -0.79, 0.51, 0.09, 0.13, 0.28, 0.14, 0.37, 0.6)
y<-c(4, 0, 2, 0, 0, 4, 2, 0, 0, 0, 4, 0, 0, 3, 0, 0, 0, 0, 0, 1 )

X<-cbind(1,x)

##  log-likelihood function

log_like<-function(beta)
{
  nu<-X%*%beta%>%exp()
  dpois(y,nu,log=TRUE)%>%sum()
}

bayes_pois<-function(x,y,init,n)
{
  
  ##  Initialise beta

  beta<-matrix(NA,ncol = 2, nrow = n)
  beta[1,]<-init

  ##  Set up MCMC

  beta_cov<-crossprod(X)%>%solve()
  beta_sd<-1

  n.accept<-0

  set.seed(14061972)

  for(i in 2:n)
  {
    bc<-rmvnorm(1,beta[i-1,],beta_sd*beta_cov)%>%t()
    alpha<-(log_like(bc)-log_like(beta[i-1,]))%>%exp()
    u<-runif(1)    
    if(u<alpha)
    {
      beta[i,]<-bc
      n.accept%<>%+(1)
    }
    else
    {
      beta[i,]<-beta[i-1,]
    }
    accept.rate<-n.accept/i
  
    ##  Automatically tune the candidate distribution
  
    if(accept.rate<0.2|accept.rate>0.4)
    {
      beta_sd%<>%multiply_by(0.7+accept.rate)
    }
  }
  return(list(beta = beta, accept.rate = accept.rate))
}

##  Now run multiple chains

N<-10000

beta0<-numeric()
beta1<-numeric()

init<-rbind(c(-1,-1),c(0,0),c(1,1))

for(j in 1:3)
{
  out<-bayes_pois(x,y,init[j,],N)
  beta0%<>%c(out$beta[,1])
  beta1%<>%c(out$beta[,2])
}

##  Plot the results

iteration<-rep(1:N,3)
chain<-c(rep("1",N),rep("2",N),rep("3",N))
name<-c(rep("beta[0]",3*N),rep("beta[1]",3*N))

df<-tibble(beta=c(beta0,beta1),iteration = rep(iteration,2),chain = rep(chain,2),name)

##  Cummulative Mean Plots

ggplot(filter(df,chain=="1")%>%group_by(name)%>%summarise(beta=cummean(beta), iteration = iteration), aes(x = iteration, y = beta, group = name))+
  geom_line()+
  facet_wrap(~name, labeller = label_parsed)

##  BGR Plot

ggplot(df,aes(x=iteration, y = beta, group = chain, color = chain ))+
  geom_line()+
  facet_wrap(~name, labeller = label_parsed)



##  Compute BGR for beta_0

S<-aggregate(beta~chain,FUN = var, data = filter(df, name =="beta[0]"))
W<-mean(S$beta)
B<-var(S$beta)
VAR<-(1-1/N)*W+B
Rhat0<-(VAR/W)%>%sqrt()

##  Compute BGR for beta_1

S<-aggregate(beta~chain,FUN = var, data = filter(df, name =="beta[1]"))
W<-mean(S$beta)
B<-var(S$beta)
VAR<-(1-1/N)*W+B
Rhat1<-(VAR/W)%>%sqrt()

```

The Brooks-Gelman-Rubin potential scale reduction factor for $\beta_0$ is $\hat{R}=$ `r round(Rhat0,6)`.\
The Brooks-Gelman-Rubin potential scale reduction factor for $\beta_1$ is $\hat{R}=$ `r round(Rhat1,6)`.