
Consider a data from a Poisson process (i.e.\ realisations of a Poisson random variable $X$) where
$$
Pr(X=x|\lambda)=\frac{\lambda^xe^{-\lambda}
}{x!}
$$
Assume a prior for $\lambda$
$$
\pi(\lambda)=4\lambda e^{-2\lambda}.
$$
For the data $\mathbf{x}=(x_1,x_2,\ldots,x_n)$ i.e.\
```{r solu0}
x<-c(2, 1, 4, 3, 4, 2, 1, 2, 0, 0)


```

a. Find the posterior distribution of $\lambda|\mathbf{x}$

For $\sum x_i=19$ and $n=10$The posterior distribution is 
$$
\pi(\lambda|\mathbf{y})=\frac{12^{21}\lambda^{20}e^{-12\lambda}}{\Gamma(21)}
$$


b. Find the $95\%$ credible interval $C$ for $\lambda$ and $\lambda|\mathbf{x}$
```{r tq1}
x<-c(2, 1, 4, 3, 4, 2, 1, 2, 0, 0)

sumx<-sum(x)
n<-length(x)

alpha<-2+sumx
beta<-2+n

##  Just compute the equal tail credible interval

ll<-qgamma(0.025,shape = alpha,rate = beta)%>%round(.,2)
ul<-qgamma(0.975,shape = alpha, rate = beta)%>%round(.,2)

```

The credible interval $C$ is (`r ll`,`r ul`)

c. Find
$$
Pr(\lambda)>1\text{ and } Pr(\lambda>1|\mathbf{x})
$$

```{r tq2}
x<-c(2, 1, 4, 3, 4, 2, 1, 2, 0, 0)

sumx<-sum(x)
n<-length(x)

alpha<-2+sumx
beta<-2+n

##  Just compute the equal tail credible interval

prior_prob<-pgamma(1,shape = 2,rate = 2)%>%round(.,2)
post_prob<-pgamma(1,shape = alpha, rate = beta)%>%round(.2)

```

$$
\begin{align}
Pr(\lambda<1)&=`r prior_prob`\\
Pr(\lambda<1|\mathbf{x})&=`r post_prob`
\end{align}
$$

d. Plot the prior and posterior of $\lambda$ together, comment?  How did the data effect the posterior compared to the prior.  
Repeat this for 
$$
\pi(\lambda)=e^{-\lambda}
$$

```{r tq3}
x<-c(2, 1, 4, 3, 4, 2, 1, 2, 0, 0)

sumx<-sum(x)
n<-length(x)

alpha<-2+sumx
beta<-2+n

x<-seq(0,5,len=1000)
prior_2<-dgamma(x,shape = 2, rate = 2)
prior_1<-dgamma(x, shape = 1 , rate = 1)
posterior_2<-dgamma(x,shape = alpha, rate = beta)
posterior_1<-dgamma(x,shape = alpha -1, rate = beta-1)

df<-tibble(X=rep(x,4),pi = c(prior_2,posterior_2,prior_1,posterior_1),names=c(rep("prior 2",1000),rep("posterior 2",1000),rep("prior 1",1000),rep("posterior 1",1000)))

ggplot(df,aes(x=X,y = pi,color=names))+
  geom_line()

```