Given 
```{r echo=FALSE}
set.seed(14061972)

x<-runif(20,-1,1)

lambda<-exp(3.1*x-1.7)

y<-rnbinom(20,mu=lambda,size = 0.7)

```
$y=$ `r round(y,2)`  and \
$x =$ `r round(x,2)`

(a) Fit a model assuming that $y_i\sim Pois(\lambda_i)$, $\log(\lambda_i)=\beta_0+\beta_1x_i$
(b) Fit a model assuming that $y_i\sim NegBinom(\mu_i,\eta)$, $\log(\mu_i)=\beta_0+\beta_1x_i$
For each model: 
(c)   Plot the trace of the parameters; 
(d)   plot the cummulative mean of the parameters and calculate the BGR for three chains; 
(e)   plot the ACF and compute the effective sample size for each parameter; and 
(f)   compute the $p_D$, $p_V$, $\bar{D}$ and $DIC$ for each of these models and compare the results.

```{r, warning=FALSE,message=FALSE}
library(mvtnorm)

x<-c(0.59, -0.67, 0.7, -0.86, -0.99, 0.92, 0.57, -0.04, -0.57, -0.74, 0.98, -0.81, -0.79, 0.51, 0.09, 0.13, 0.28, 0.14, 0.37, 0.6)
y<-c(6, 0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 0, 1, 0, 1, 0, 1, 1, 0)

##  log-likelihood function

bayes_pois<-function(x,y,init,n)
{

set.seed(14061972)
  
  X<-cbind(1,x)
  
  log_like<-function(beta)
    {
      nu<-X%*%beta%>%exp()
      dpois(y,nu,log=TRUE)%>%sum()
    }
  
  ##  Initialise beta

  beta<-matrix(NA,ncol = 2, nrow = n)
  beta[1,]<-init
  dev<-numeric(length = n)
  dev[1]<--2*log_like(init)

  ##  Set up MCMC

  beta_cov<-crossprod(X)%>%solve()
  beta_sd<-1

  n.accept<-0

  set.seed(14061972)

  for(i in 2:n)
  {
    bc<-rmvnorm(1,beta[i-1,],beta_sd*beta_cov)%>%t()
    alpha<-(log_like(bc)-log_like(beta[i-1,]))%>%exp()
    u<-runif(1)    
    if(u<alpha)
    {
      beta[i,]<-bc
      n.accept%<>%+(1)
    }
    else
    {
      beta[i,]<-beta[i-1,]
    }
    accept.rate<-n.accept/i
  
    ##  Automatically tune the candidate distribution
  
    if(accept.rate<0.2|accept.rate>0.4)
    {
      beta_sd%<>%multiply_by(0.7+accept.rate)
    }
    dev[i]<--2*log_like(beta[i,])
  }
  pD<-mean(dev)+2*log_like(colMeans(beta))
  pV<-var(dev)/2
  Dbar<-mean(dev)
  DICd<-Dbar+pD
  DICv<-Dbar+pV
  return(list(beta = beta, accept.rate = accept.rate, Dbar = Dbar, DICd = DICd, DICv = DICv, pD = pD, pV= pV))
}

#####################

bayes_nbinom<-function(x,y,init,n)
{
  
set.seed(14061972)
  
  log_like<-function(BETA,ETA)
  {
    mu<-X%*%BETA%>%exp()
    dnbinom(y,size = ETA,mu = mu, log = TRUE)%>%sum()
  }
  
  X<-cbind(1,x)
  
  beta<-matrix(NA,ncol = 2, nrow = n)
  eta<-numeric(length = n)
  dev<-numeric(length = n)
  dev[1]<-dnbinom(y,size = init[3],mu = exp(X%*%init[1:2]),log = TRUE)%>%sum()%>%multiply_by(-2)
  beta[1,]<-init[1:2]
  eta[1]<-init[3]
  beta_cov<-crossprod(X)%>%solve()
  beta_sd<-1.5
  eta_var<-1
  beta.accept<-0
  eta.accept<-0
  
    for(i in 2:n)
  {
    bc<-rmvnorm(1,mean = beta[i-1,],sigma = beta_sd*beta_cov)%>%t()
    alpha<-(log_like(bc,eta[i-1])-log_like(beta[i-1,],eta[i-1]))%>%exp()
    u<-runif(1)
    if(alpha>u)
    {
      beta[i,]<-bc
      beta.accept%<>%+(1)
    }
    else
    {
      beta[i,]<-beta[i-1,]
    }
    beta.accept.rate<-beta.accept/i
        
    ##  Automatically tune the candidate distribution
  
    if(beta.accept.rate<0.2|beta.accept.rate>0.4)
    {
      beta_sd%<>%multiply_by(0.7+beta.accept.rate)
    }
    

  ##  update eta

    ac<-eta[i-1]^2/eta_var
    bc<-eta[i-1]/eta_var

    ec<-rgamma(1,ac,bc)

    a<-ec^2/eta_var
    b<-ec/eta_var

    alpha<-(log_like(beta[i,],ec)-log_like(beta[i,],eta[i-1])+dgamma(eta[i-1],a,b,log=TRUE)-dgamma(ec,ac,bc,log = TRUE)-ec+eta[i-1])%>%exp()

  u<-runif(1)
  if(alpha>u)
  {
    eta[i]<-ec
    eta.accept%<>%+(1)
  }
  else
  {
    eta[i]<-eta[i-1]
  }

    ##  Automatically tune the candidate distribution
  
    eta.accept.rate<-eta.accept/i
  
    if(eta.accept.rate<0.2|eta.accept.rate>0.4)
    {
      eta_var%<>%multiply_by(0.7+eta.accept.rate)
    }
  
    dev[i]<--2*log_like(beta[i,],eta[i])
    }
  pD<-mean(dev)+2*log_like(colMeans(beta),mean(eta))
  pV<-var(dev)/2
  Dbar<-mean(dev)
  DICd<-Dbar+pD
  DICv<-Dbar+pV
  return(list(beta = beta, eta = eta, beta.accept = beta.accept, beta.accept.rate = beta.accept.rate,eta.accept = eta.accept, eta.accept.rate = eta.accept.rate, Dbar = Dbar, DICd = DICd, DICv = DICv, pD = pD, pV= pV))
}

##  Now run multiple chains

N<-10000

pois_out<-list()
nb_out<-list()

init<-rbind(c(-1.7,3.1,1),c(0,1,2),c(0,0,3))


for(j in 1:3)
{
  pois_out[[j]]<-bayes_pois(x,y,init[j,1:2],N)
  nb_out[[j]]<-bayes_nbinom(x,y,init[j,],N)
}

p.Dbar<-pois_out[[1]]$Dbar
p.pD<-pois_out[[1]]$pD
p.pV<-pois_out[[1]]$pV
p.DICd<-pois_out[[1]]$DICd
p.DICv<-pois_out[[1]]$DICv

nb.Dbar<-nb_out[[1]]$Dbar
nb.pD<-nb_out[[1]]$pD
nb.pV<-nb_out[[1]]$pV
nb.DICd<-nb_out[[1]]$DICd
nb.DICv<-nb_out[[1]]$DICv

##  Plot the results for Poisson

iteration<-rep(1:N,3)
chain<-c(rep("1",N),rep("2",N),rep("3",N))
name<-c(rep("beta[0]",3*N),rep("beta[1]",3*N))

beta.pois<-rbind(pois_out[[1]]$beta,pois_out[[2]]$beta,pois_out[[3]]$beta)

df.pois<-tibble(beta=c(beta.pois),iteration = rep(iteration,2),chain = rep(chain,2),name)

##  Trace Plots

ggplot(filter(df.pois,chain =="1"), aes(x= iteration, y = beta))+
  geom_line()+
  facet_wrap(~name, labeller = label_parsed)

##  Cummulative Mean Plots

ggplot(filter(df.pois,chain=="1")%>%group_by(name)%>%summarise(beta=cummean(beta), iteration = iteration), aes(x = iteration, y = beta, group = name))+
  geom_line()+
  facet_wrap(~name, labeller = label_parsed)

##  BGR Plot

ggplot(df.pois,aes(x=iteration, y = beta, group = chain, color = chain ))+
  geom_line()+
  facet_wrap(~name, labeller = label_parsed)

##  Plot the results for Negative Binomial

iteration<-rep(1:N,3)
chain<-c(rep("1",N),rep("2",N),rep("3",N))
name<-c(rep("beta[0]",3*N),rep("beta[1]",3*N),rep("eta",3*N))

beta.nb<-rbind(nb_out[[1]]$beta,nb_out[[2]]$beta,nb_out[[3]]$beta)
eta<-c(nb_out[[1]]$eta,nb_out[[2]]$eta, nb_out[[3]]$eta)

df.nb<-tibble(beta=c(beta.nb,eta),iteration = rep(iteration,3),chain = rep(chain,3),name)

##  Trace Plots

ggplot(filter(df.nb,chain =="1"), aes(x= iteration, y = beta))+
  geom_line()+
  facet_wrap(~name, labeller = label_parsed)

##  Cummulative Mean Plots

ggplot(filter(df.nb,chain=="1")%>%group_by(name)%>%summarise(beta=cummean(beta), iteration = iteration), aes(x = iteration, y = beta, group = name))+
  geom_line()+
  facet_wrap(~name, labeller = label_parsed)

##  BGR Plot

ggplot(df.nb,aes(x=iteration, y = beta, group = chain, color = chain ))+
  geom_line()+
  facet_wrap(~name, labeller = label_parsed)

##  BGR Computations

BGR<-function(par,DF)
{
  S<-aggregate(beta~chain,FUN = var, data = filter(DF, name == par))
  W<-mean(S$beta)
  B<-var(S$beta)
  VAR<-(1-1/N)*W+B
  Rhat<-(VAR/W)%>%sqrt()
  return(Rhat)
}

bgr.pois<-sapply(unique(df.pois$name),function(x) BGR(x, df.pois))
bgr.nb<-sapply(unique(df.nb$name),function(x) BGR(x, df.nb))

##  ACF and ESS

##  For Poisson

acf_beta_0<-acf(pois_out[[1]]$beta[,1],plot = FALSE)$acf
acf_beta_1<-acf(pois_out[[1]]$beta[,2],plot = FALSE)$acf
lag<-rep(1:length(acf_beta_0),2)
name<-c(rep("beta[0]",length(acf_beta_0)),rep("beta[1]",length(acf_beta_1)))

df<-tibble(acf = c(acf_beta_0,acf_beta_1),lag,name)

ggplot(df,aes(x = lag, y = acf))+
  geom_bar(stat = "identity",alpha = 0.5)+
  geom_hline(yintercept = 0.05,linetype = "dashed")+
  facet_wrap(~name,labeller = label_parsed)

##  For negative-binomial

acf_beta_0<-acf(nb_out[[1]]$beta[,1],plot = FALSE)$acf
acf_beta_1<-acf(nb_out[[1]]$beta[,2],plot = FALSE)$acf
acf_eta<-acf(nb_out[[1]]$eta,plot = FALSE)$acf
lag<-rep(1:length(acf_beta_0),3)
name<-c(rep("beta[0]",length(acf_beta_0)),rep("beta[1]",length(acf_beta_1)),rep("eta",length(acf_eta)))

df<-tibble(acf = c(acf_beta_0,acf_beta_1,acf_eta),lag,name)

ggplot(df,aes(x = lag, y = acf))+
  geom_bar(stat = "identity",alpha = 0.5)+
  geom_hline(yintercept = 0.05,linetype = "dashed")+
  facet_wrap(~name,labeller = label_parsed)

pois.ess<-mcmcse::ess(pois_out[[1]]$beta)
nb.ess<-mcmcse::ess(cbind(nb_out[[1]]$beta,nb_out[[1]]$eta))

```

###   Brooks-Gelman-Rubin Potential Scale Reductions Factors

|  Model     |  $\beta_0$              |  $\beta_1$              |  $\eta$               |
|------------|-------------------------|-------------------------|-----------------------|
|  Poisson   | `r round(bgr.pois[1],3)`| `r round(bgr.pois[2],3)`|  -                    |
|  Neg-Binom | `r round(bgr.nb[1],3)`  |  `r round(bgr.nb[2],3)` | `r round(bgr.nb[3],3)`|

###   Effective Sample Sizes

|  Model     |  $\beta_0$              |  $\beta_1$              |  $\eta$               |
|------------|-------------------------|-------------------------|-----------------------|
|  Poisson   | `r round(pois.ess[1],3)`| `r round(pois.ess[2],3)`|  -                    |
|  Neg-Binom | `r round(nb.ess[1],3)`  |  `r round(nb.ess[2],3)` | `r round(nb.ess[3],3)`|

###   Deviance Information Criteria

|Model      | $\bar{D}$  |$p_D$    |$p_v$     |$DIC_{p_D}$ |$DIC_{p_V}$ |
|-----------|------------|---------|----------|------------|------------|
|Poisson    | `r p.Dbar` |`r p.pD` | `r p.pV` |`r p.DICd`  |`r p.DICv`  |
| Neg-Binom | `r nb.Dbar`|`r nb.pD`|`r nb.pV` |`r nb.DICd` |`r nb.DICv` |

